\pkg{Rgbp}: An R Package for Hierarchical Modeling and Method Checking

### 2013-07-10 jss/article.tex
paper
editting paper

%% an abstract and keywords
\Abstract{
\pkg{Rgbp} is an R package for estimating and validating a two-level model (a random-effects mixed model). The estimation procedure utilizes Bayesian machinery and the validation involves checking frequency properties of the procedure via repeated sampling (which we call ``method checking"). It is found that \pkg{Rgb} characterized by such Bayesian-frequentist reconciliation has strong points of both sides, working well in small samples and yielding good coverage probability. Following three examples for Gaussian, Poisson and Binomial data validate this point.}
\Keywords{hierarchical model, multilevel model, random effects mixed model, method checking, coverage probability, Normal, Poisson, Binomial,  shrinkage, \proglang{R}}
\Plainkeywords{keywords, comma-separated, not capitalized, r} %% without formatting
%% at least one keyword must be supplied

@@

Once we estimate these two moments of shrinkage, we can also estimate the posterior moments given only the data ($\textbf{y}$). Taking the Normal-Normal model as an example we note the following identities
\begin{eqnarray}
E(\mu_{j}\vert \textbf{y}) & = & E(E(\mu_{j}\vert \textbf{y}, \beta, A)\vert \textbf{y}) =E((1-B_{j})y_{j}+B_{j}\mu_{0j}\vert \textbf{y}).\\ 
Var(\mu_{j}\vert \textbf{y}) & = & E(Var(\mu_{j}\vert \textbf{y}, \beta, A)\vert \textbf{y})+Var(E(\mu_{j}\vert \textbf{y}, \beta, A)\vert \textbf{y}).
\end{eqnarray}  
Note that both $E(\mu_{j}\vert \textbf{y}, \beta, A)$ and $Var(\mu_{j}\vert \textbf{y}, \beta, A)$ are linear functions of the shrinkage factors as specified in (\ref{normalpost}) and since we can estimate $E(B_{j}\vert\textbf{y})$ and $Var(B_{j}\vert\textbf{y})$ via ADM, we can finally estimate $E(\mu_{j}\vert \textbf{y})$ and $Var(\mu_{j}\vert \textbf{y})$.

@@

The output contains information about sample mean ($y_{j}$), caseload ($n_{j}$), known prior mean ($\lambda_{0}$), shrinkage estimate ($\hat{B}_{j}$), lower bound of interval estimate ($\hat{\lambda}_{low}$), posterior mean ($\hat{\lambda}=E(\lambda_{j}\vert \textbf{z})$), upper bound of interval estimate ($\hat{\lambda}_{upp}$), and standard deviation of posterior distribution ($sd(\lambda_{j}\vert \textbf{z})$).

@@

The regression towards the mean (RTTM) is obvious in the left-side graph; the observed sample means, empty dots on the upper line, are shrinking towards the known second-level mean (a blue vertical line at 0.02) to the different extents. Note that some hospitals' ranks have changed by shrinking much harder towards 0.02 than others. For example, the empty square symbol at the crossing point of the two left-most lines (8th and 23rd hospitals on the list above) indicates that seemingly safest hospital among 31 hospitals in terms of the observed sample mean was not actually safer than the second safest hospital.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> data(schools)
R> attach(schools)
R> g <- gbp(y, se, model = "gr")
R> g
\end{CodeInput}
\begin{CodeOutput}
Summary for each unit (sorted by n):

@@

Note that the posterior standard deviation tends to be bigger among outfielders than among others. Let's see its formula in (9). We can notice that the estimated posterior mean is a critical factor that makes posterior variances different from each player because every player has the same at-bats ($n_{j}$) and $r$. As the posterior mean gets closer to 0.5, the posterior variance gets bigger and has the biggest value when the posterior mean is 0.5. Intuitively, it makes sense. If a true batting average is 0.1, we can easily expect less hits. But if it is 0.5, we are less certain whether an at-bat would be more likely a hit or not, like a coin tossing. Our model is automatically reflecting on such uncertainty.
\begin{CodeChunk}
\begin{CodeInput}
R> summary(b)
\end{CodeInput}
\begin{CodeOutput}
Main summary:

@@

This summary also includes the result of logistic regression fit because we did not know prior means in advance. We had to estimate them via this logistic regression model. Let's look at the p-value of the indicator covariate. It shows that the two prior means distinguishing outfielders from other positions were significantly different (p-value = 0.038), supporting our initial belief. The positive sign of $\hat{\beta}_{1}$ indicates that the mean batting average for all the outfielders tends to be higher than that for those in the other positions.
\\
\begin{CodeChunk}
\begin{CodeInput}
R> plot(b)
\end{CodeInput}
\end{CodeChunk}
\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.3]{baseball1.png}
\caption{Shrinkage plot and 95\% interval plot}
\end{center}
\end{figure}

Let's see shrinkage and 95\% interval plots for more intuition. It is obvious that sample means (empty dots) on the upper line shrink towards two different means, 0.310 and 0.233. For reference, the red line symbols on dots are for when two or more points have the same mean and are plotted over each other. For example, five players (from the 11th player to the 15th) have the same sample mean (0.222) and at this point on the upper line, there are short red lines toward five directions.

For reference, the observed sample mean among outfielders is 0.308 and that among other positions is 0.231. But why are the estimated two prior means from this Binomial model 0.310 and 0.233 each? This can be attributed to the logistic regression which estimates prior mean by a non-linear logit function of $x'\beta$, causing a small bias (see 3.3). The Normal model can avoid this small bias because it uses a linear regression to estimate these prior means.

The 95\% interval plot shows us range of true batting average for each player, which clarifies the regression toward the mean (RTTM) within two groups. Let's see the 10th, 14th, and 15th players on the graph for example. They are outfielders but their batting averages (sample mean) are far lower than the first five outfielders. RTTM means that it can happen by their bad luck though in the long run their batting averages will come back to their expected ones as outfielders (blue horizontal line). The fact that their sample means are closer to the lower part of their 95\% intervals supports this argument. So, their posterior means (red dots) can also be interpreted as their RTTM.

Then, how much can we trust these interval estimates? The following method checking that regards the estimated values, $\hat{r}$ (=112.95) and $\hat{\beta}~(=(-1.194, ~0.389)^{T})$, as true values when it generates pseudo-datasets will answer it.

>>>>

%% an abstract and keywords
\Abstract{
\pkg{Rgbp} is an R package for estimating and validating a two-level model (a random-effects mixed model). The estimation procedure utilizes Bayesian machinery and the validation involves checking frequency properties of the procedure via repeated sampling (which we call ``method checking"). It is found that \pkg{Rgb} characterized by such Bayesian-frequentist reconciliation has strong points of both sides, working well in small samples and yielding good coverage probability. Following three examples for Gaussian, Poisson and Binomial data will validate this point.}
\Keywords{hierarchical model, multilevel model, random effects mixed model, method checking, coverage probability, Normal, Poisson, Binomial,  shrinkage, \proglang{R}}
\Plainkeywords{keywords, comma-separated, not capitalized, r} %% without formatting
%% at least one keyword must be supplied

@@

Once we estimate these two moments of shrinkage, we can also estimate the posterior moments given only the data ($\textbf{y}$). Taking the Normal-Normal model as an example we note the following identities
\begin{eqnarray}
E(\mu_{j}\vert \textbf{y}) & = & E(E(\mu_{j}\vert \textbf{y}, \beta, A)\vert \textbf{y}), \\
Var(\mu_{j}\vert \textbf{y}) & = & E(Var(\mu_{j}\vert \textbf{y}, \beta, A)\vert \textbf{y})+Var(E(\mu_{j}\vert \textbf{y}, \beta, A)\vert \textbf{y}).
\end{eqnarray}  
Note that both $E(\mu_{j}\vert \textbf{y}, \beta, A)$ and $Var(\mu_{j}\vert \textbf{y}, \beta, A)$ are linear functions of the shrinkage factors as specified in (\ref{normalpost}) and since we can estimate $E(B_{j}\vert\textbf{y})$ and $Var(B_{j}\vert\textbf{y})$ via ADM, we can finally estimate $E(\mu_{j}\vert \textbf{y})$ and $Var(\mu_{j}\vert \textbf{y})$.

@@

The output contains information about sample mean ($y_{j}$), caseload ($n_{j}$), known prior mean ($\lambda_{0}$), shrinkage estimate ($\hat{B}_{j}$), lower bound of interval estimate ($\hat{\lambda}_{low}$), posterior mean ($\hat{\lambda}=E(\lambda_{j}\vert \textbf{y})$), upper bound of interval estimate ($\hat{\lambda}_{upp}$), and standard deviation of posterior distribution ($sd(\lambda_{j}\vert \textbf{y})$).

@@

The regression towards the mean (RTTM) is obvious in the left-side graph; the observed sample means, empty dots on the upper horizontal line, are shrinking towards the known second-level mean (a blue vertical line at 0.02) to the different extents. Note that some hospitals' ranks have changed by shrinking much harder towards 0.02 than others. For example, the empty square symbol at the crossing point of the two left-most lines (8th and 23rd hospitals on the list above) indicates that seemingly the safest hospital among 31 hospitals in terms of the observed death ratio was not actually safer than seemingly the second safest hospital.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> data(schools)
R> attach(schools)
R> g <- gbp(y, se, model = "gr")
R> g
\end{CodeInput}
\begin{CodeOutput}
Summary for each unit (sorted by se):

@@

%Note that the posterior standard deviation tends to be bigger among outfielders than among others. Let's see its formula in (9). We can notice that the estimated posterior mean is a critical factor that makes posterior variances different from each player because every player has the same at-bats ($n_{j}$) and $r$. As the posterior mean gets closer to 0.5, the posterior variance gets bigger and has the biggest value when the posterior mean is 0.5. Intuitively, it makes sense. If a true batting average is 0.1, we can easily expect less hits. But if it is 0.5, we are less certain whether an at-bat would be more likely a hit or not, like a coin tossing. Our model is automatically reflecting on such uncertainty.

The below summary includes the result of logistic regression fit because we did not know prior means in advance. We had to estimate them via this logistic regression model. Let's look at the p-value of the indicator covariate. It shows that the two prior means distinguishing outfielders from other positions were significantly different (p-value = 0.038), supporting our initial belief. The positive sign of $\hat{\beta}_{1}$ indicates that the mean batting average for all the outfielders tends to be higher than that for those in the other positions.

\begin{CodeChunk}
\begin{CodeInput}
R> summary(b)
\end{CodeInput}
\begin{CodeOutput}
Main summary:

@@

Now, let's see following shrinkage and 95\% interval plots for more intuition. It is obvious that sample means (empty dots) on the upper line shrink towards two different means, 0.310 and 0.233. For reference, the red line symbols on dots are for when two or more points have the same mean and are plotted over each other. For example, five players (from the 11th player to the 15th) have the same sample mean (0.222) and at this point on the upper horizontal line, there are short red lines toward five directions.

\begin{CodeChunk}
\begin{CodeInput}
R> plot(b)
\end{CodeInput}
\end{CodeChunk}
\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.3]{baseball1.png}
\caption{Shrinkage plot and 95\% interval plot}
\end{center}
\end{figure}

%For reference, the observed sample mean among outfielders is 0.308 and that among other positions is 0.231. But why are the estimated two prior means from this Binomial model 0.310 and 0.233 each? This can be attributed to the logistic regression which estimates prior mean by a non-linear logit function of $x'\beta$, causing a small bias (see 3.3). The Normal model can avoid this small bias because it uses a linear regression to estimate these prior means.

The 95\% interval plot shows us range of true batting average for each player, which clarifies the regression toward the mean (RTTM) within two groups. Let's see the 10th, 14th, and 15th players on the graph for example. They are outfielders but their observed batting averages are far lower than the first five outfielders. RTTM means that it can happen by their bad luck though in the long run their batting averages will come back to their expected ones as outfielders (blue horizontal line). The fact that their sample means are closer to the lower part of their 95\% intervals supports this argument. So, their posterior means (red dots) can also be interpreted as their RTTM.

Then, how much can we trust these interval estimates? The following method checking that regards the estimated values, $\hat{r}$ (=112.95) and $\hat{\beta}~(=(-1.194, ~0.389)^{T})$, as true values when it generates pseudo-datasets will answer it. For reference, if we want to try different true parameters, the appropriate code will be \code{coverage(b, A.or.r = 100, reg.coef = c(-1, 0.2), nsim = 10000)} instead of below one.
